---
title: "Assignment 1 ST464/ST644"
author: "Elizabeth O'Dwyer 21253265"
date: "`r format(Sys.time(), '%X %d %B, %Y')`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width=20, fig.height=12)
```

```{r, eval=T, echo=FALSE}
suppressMessages(library(ggplot2))
suppressMessages(library(ggdendro))
suppressMessages(library(tidyverse))
suppressMessages(library(ggplot2))
suppressMessages(library(GGally))
suppressMessages(library(dendextend))

```



#### Question 4
```{r eval=T} 
# change to eval=T
w <- read.csv(here::here("hwk", "Data","worldhappiness2019.csv"))
rownames(w)<- w$Country
names(w)<- c( "Rank", "Country", "Score", "GDP", "Family", "Life", "Freedom", "Generosity", "Trust")
rownames(w)<- w$Country
wscores <- w[,-(1:3)]
```

(a)
```{r}
cor_matr <- cor(wscores, method="pearson")

#Highest correlation is between GDP and Life.
w_df = data.frame(wscores)
b <- ggplot(w_df, aes(x = Life, y = GDP))
b + geom_point()
```
The outlier of this graph has a low 'Life' value.
To find this I'll retrieve the minimum value from the
Life column.
```{r}
index_outlier <- which(w_df$Life == min(w_df$Life))
outlier_country <- row.names(w_df)[index_outlier]
print(paste('Outlier Country:',outlier_country, sep=' '))
```


(b)
```{r}
D1 <- as.matrix(dist(wscores,method="euclidean"))
h<- hclust(as.dist(D1),"single")
par(mar=c(1, 1, 1, 1))
dend <- as.dendrogram(h)
dend_withcolor <- color_branches(dend,k=7, col=c(2,3,5,7,9,10,11))
plot(dend_withcolor)
```

Outliers are the countries on the left of the plot that are not a part of the huge green cluster.
Outliers consist of African Republic, Rwanda, Syria, Somalia, Swaziland and Burundi.
They are all isolated clusters in the 7 cluster solution.



(c)
```{r}
h_3 <- unname(cutree(h,3))
table(h_3)
```
In cluster1 there are 154 values, in cluster2 there is one value and in cluster3 there is one value.
```{r}
w_df$h_clusters <- as.factor(h_3)
print(w_df['Ireland','h_clusters'])
print(w_df['India','h_clusters'])
```

```{r}
source('~/ST464/sumPartition.R')
sp <- sumPartition(wscores,h_3)
```
Cluster 1 has the highest Family score and overall scores.



(d)
```{r}
ggparcoord(w_df, columns=1:6, groupColumn="h_clusters",scale="uniminmax")+
xlab("")+ ylab("Score")
```
The lowest-scoring cluster from this graph has low GDP, Family,Life and Trust scores but high Freedom and Generosity.




(e)
```{r}
set.seed(123)
km <- kmeans(wscores, 3,nstart=10)
table(km$cluster)
w_df$km_clusters <- km$cluster
```

Look for Outliers (arbitrarily looking for the top 5 furthest from centre.).
```{r}
centers <- km$centers[km$cluster, ]
distances <- sqrt(rowSums((wscores - centers)^2))
outliers <- order(distances, decreasing=T)[1:5]
print(wscores[outliers,])

```
(f)
```{r}
ggparcoord(w_df, columns=1:6, groupColumn='km_clusters',scale="uniminmax")+
xlab("")+ ylab("Score")
```
Cluster1 has the highest family and score overall.


 
(g)
```{r}
w$km <- km$cluster
boxplot(Score ~ km, data = w,xlab = "Cluster",
   ylab = "Score", main = "Box Plot of Score for K_means Clusters",names = c('Cluster 1', 'Cluster 2', 'Cluster 3'))
```

Cluster 1 has the highest median. The cluster has most of the data occuring within the first quartile.

The second cluster has the second highest median. The data has a bigger range than the first cluster. The median occurs roughly midway between the first and 3rd quartile.

The third cluster has the lowest median. The range is similar in size to cluster 2. The majority of the data occurs close to the third quartile.




